---
title: "Linear Regression"
subtitle: "UCLA Soc 114"
format:
  revealjs: default # For the web slides
  #beamer: default      # To generate the PDF version
execute:
  echo: true
resources: 
   - lec05_linear_regression.pdf
---

```{r, echo = FALSE, message = FALSE, warning = FALSE}
library(tidyverse)
library(scales)
library(foreach)
options(pillar.print_max = 3, pillar.print_min = 3)
theme_set(theme_classic())
update_theme(text = element_text(size = 18))
set.seed(90095)
```

## Linear regression: Learning goals

Some things you may know

- How to fit a linear model
- How to make predictions

Data science ideas

- Why model at all?
- Penalized linear regression

## Data for illustration

U.S. adult income by

- sex (male, female)
- age (30--50)
- year (2010--2019)

among those working 35+ hours per week for 50+ weeks per year. Data are simulated based on the 2010–2019 American Community Survey (ACS).

## Data for illustration

The function below will simulate data

```{r}
simulate <- function(n = 100) {
  read_csv("https://ilundberg.github.io/description/assets/truth.csv") |>
    slice_sample(n = n, weight_by = weight, replace = T) |>
    mutate(income = exp(rnorm(n(), meanlog, sdlog))) |>
    select(year, age, sex, income)
}
```

## Data for illustration

```{r}
simulated <- simulate(n = 3e4)
```
```{r, echo = F}
simulated
```

## Conditional expectation

. . .

Mean of an outcome within a population subgroup.

. . .

- **expectation** refers to taking a mean

. . .

- **conditional** refers to within a subgroup

. . .

Example: Mean income among females age 47 in 2019 

. . .

**Task.** Estimate this in our data.

## Code: Find the subgroup

`filter()` restricts our data to cases meeting requirements:

- the `sex` variable equals the value `female`
- the `age` variable equals the value `47`
- the `year` variable equals the value `2019`

```{r}
subgroup <- simulated |>
  filter(sex == "female") |>
  filter(age == 47) |>
  filter(year == 2019)
```

## Code: Estimate the mean

`summarize()` aggregates to the mean

```{r}
subgroup |>
  summarize(conditional_expectation = mean(income))
```

## Code: Mean in many subgroups

. . .

With `group_by`, you can `summarize` many subgroups

```{r}
simulated |>
  group_by(sex, age, year) |>
  summarize(conditional_expectation = mean(income))
```

## Conditional expectation: Math

. . .

The **conditional expectation function** is the subgroup mean of $Y$ within a subgroup with the predictor values $\vec{X} = \vec{x}$.

$$
f(\vec{x}) = \text{E}(Y\mid\vec{X} = \vec{x})
$$

To learn $f(\vec{x})$ from data is a central task in **statistical learning**.

# Statistical Learning by Pooling Information

## A subgroup is small

. . .

```{r}
simulated |>
  filter(sex == "female") |>
  filter(year == 2019) |>
  filter(age == 47)
```

. . .

Very few cases $\rightarrow$ statistically uncertain 

. . .

**How to better estimate for 47-year-old females in 2019?**

## Pooling information across subgroups

. . .

We have many female respondents in 2019. Few are age 47.

. . .

```{r}
simulated |>
  filter(sex == "female") |>
  filter(year == 2019)
```

. . .

Could we use them to learn about the 47-year-olds?

## Pooling information across subgroups

```{r, echo = FALSE, fig.height = 4}
female_2019 <- simulated |>
  filter(sex == "female") |>
  filter(year == 2019)
fit <- lm(income ~ age, data = female_2019)
target_yhat <- predict(fit, newdata = tibble(age = 47))
p <- female_2019 |>
  mutate(yhat = predict(fit)) |>
  group_by(sex, age) |>
  summarize_all(mean) |>
  ggplot(aes(x = age, y = income)) +
  geom_point(color = "gray") +
  scale_y_continuous(
    name = "Annual Income Y",
    labels = label_currency(),
  ) +
  labs(x = "Age X")
p
```

## Pooling information across subgroups

```{r, echo = F, fig.height = 4}
p <- p +
  # Regression line
  geom_line(aes(y = yhat)) +
  annotate(
    geom = "text", x = 37, y = 50e3, 
    label = "E('Y | X = x') == beta[0] + beta[1] * x",
    parse = TRUE, size = 10
  ) 
p
```

## Pooling information across subgroups

```{r, echo = F, fig.height = 4}
p <- p +
  # Prediction
  geom_segment(
    x = 47, xend = 47, y = -Inf, yend = target_yhat, 
    color = "#2774AE", linetype = "dashed"
  ) +
  annotate(geom = "text", x = 47.5, y = 45e3, label = "X = 47", color = "#2774AE", hjust = 0, size = 10) +
  geom_point(
    x = 47, y = target_yhat, 
    color = "#2774AE", size = 3
  ) +
  geom_segment(
    x = 47, xend = -Inf, y = target_yhat, yend = target_yhat, 
    color = "#2774AE", linetype = "dashed"
  ) +
  annotate(
    geom = "text", x = 40, y = target_yhat, 
    label = "hat(Y) == ~hat(E)('Y | X' == 47) == ~hat(beta)[0] + hat(beta)[1]%*%47", 
    parse = TRUE, size = 10,
    color = "#2774AE",
    vjust = -1
  )
p
```

## Practice question

$$
\text{E}(Y\mid X) = \beta_0 + \beta_1 X
$$

Suppose $\beta_0 = 5$ and $\beta_1 = 3$

1. What is the conditional mean when $X = 0$?
2. What is the conditional mean when $X = 1$?
3. What is the conditional mean when $X = 2$?
4. How much does the conditional mean change for each unit increase in $X$?

## Code

The next slides explain how to code a model in R.

## Code: Simulate data

. . .

Generate some data
```{r}
simulated <- simulate(n = 3e4)
```

. . .

Restrict to female respondents in 2019
```{r}
female_2019 <- simulated |>
  filter(sex == "female") |>
  filter(year == 2019)
```

. . .

(Below is `simulate` if you did not copy it before)

```{r, eval = F}
simulate <- function(n = 100) {
  read_csv("https://ilundberg.github.io/description/assets/truth.csv") |>
    slice_sample(n = n, weight_by = weight, replace = T) |>
    mutate(income = exp(rnorm(n(), meanlog, sdlog))) |>
    select(year, age, sex, income)
}
```

## Code: Learn a model

```{r}
model <- lm(
  formula = income ~ age, 
  data = female_2019
)
```

- `model` is an object of class `lm` for **l**inear **m**odel
- `lm()` function creates this object
- `formula` argument is a model formula
     - `outcome ~ predictor` is the syntax
- `data` is a dataset containing `outcome` and `predictor`

## Code: Examine the learned model

```{r}
summary(model)
```

## Code: Predict for a new X value

. . .

Define X value at which to predict
```{r}
to_predict <- tibble(age = 47)
```

. . .

Predict for that subgroup
```{r}
predict(model, newdata = to_predict)
```

. . .

Recap: Our model **pooled information**:

- People of all ages contributed to `model`
- Then we predicted at a single age

## Code: Three steps

- Estimate a model
- Define $x$ to predict
- Predict $\hat{Y} = \hat{\text{E}}(Y\mid X = x)$

**What if you were going to do this many times on different data?**

## Code: Three steps in a function

. . .

```{r}
estimator <- function(data) {
  # Learn the model from the data
  model <- lm(formula = income ~ age, data = data)
  # Define our target subgroup
  to_predict <- tibble(age = 47)
  # Predict
  estimate <- predict(model, newdata = to_predict)
  # Return the estimate
  return(estimate)
}
```

## Code: All together

```{r}
estimator(data = female_2019)
```

. . .

```{r}
female_2019 |>
  estimator()
```

. . .

```{r}
simulated |> 
  filter(sex == "female") |>
  filter(year == 2010) |>
  estimator()
```

## Practice question

Below is the line fit to the population data. Suppose we want to learn $\text{E}(\log(Y)\mid X = 30)$.

```{r, echo = FALSE, fig.height = 3}
read_csv("https://ilundberg.github.io/description/assets/truth.csv") |>
  filter(sex == "female" & year == 2019) |>
  ggplot(aes(x = age, y = meanlog)) +
  geom_point() +
  geom_smooth(method = "lm", se = F) +
  labs(
    y = "Mean of log income",
    x = "Age",
    caption = "Among female respondents in 2019"
  )
```

1. Why might this model make a misleading estimate?
2. Why might the model still be useful?

# Additive vs Interactive

## Two models

```{r, echo = FALSE, fig.height = 4}
all_2019 <- simulated |>
  filter(year == 2019)
model <- lm(income ~ age*sex, data = all_2019)
target_subgroup <- tibble(age = 47, sex = c("female","male"))
target_subgroup$yhat <- predict(model, newdata = target_subgroup)

p <- all_2019 |>
  mutate(yhat = predict(model)) |>
  group_by(sex, age) |>
  summarize_all(mean) |>
  ggplot(aes(x = age, y = income)) +
  geom_point(color = "gray", size = 1) +
  scale_y_continuous(
    name = "Annual Income Y",
    labels = label_currency(),
  ) +
  labs(x = "Age X") +
  facet_wrap(~sex)
p
```

## Two models

```{r, echo = F, fig.height = 4}
p2 <- p +
  # Regression line
  geom_line(aes(y = yhat)) +
  geom_text(
    data = tibble(
      sex = c("female","male"),
      age = 30, income = 51e3,
      label = c(
        "E('Y | X = x, Sex = Female') == {beta[0]}^'Female' + {beta[1]}^'Female' * x",
        "E('Y | X = x, Sex = Male') == {beta[0]}^'Male' + {beta[1]}^'Male' * x"
      )
    ),
    aes(label = label),
    parse = T, size = 5, hjust = 0
  )
p2
```

## Two models

```{r, echo = F, fig.height = 4}
p3 <- p2 +
  # Prediction
  geom_segment(
    data = target_subgroup,
    aes(x = 47, xend = 47, y = -Inf, yend = yhat), 
    color = "#2774AE", linetype = "dashed"
  ) +
  geom_segment(
    data = target_subgroup,
    aes(x = 47, xend = -Inf, y = yhat, yend = yhat), 
    color = "#2774AE", linetype = "dashed"
  ) +
  geom_point(
    data = target_subgroup,
    aes(x = age, y = yhat), 
    color = "#2774AE", size = 3
  )
p3
```

## Two models: Interaction

$$
\begin{aligned}
\text{E}(Y\mid X, \text{Female}) &= \beta_0^\text{Female} + \beta_1^\text{Female}\times \text{Age} \\
\text{E}(Y\mid X, \text{Male}) &= \beta_0^\text{Male} + \beta_1^\text{Male}\times \text{Age} \\
\end{aligned}
$$

. . .

Equivalently,
$$\text{E}(Y \mid X, \text{Sex}) = \gamma_0 + \gamma_1(\text{Female}) + \gamma_2(\text{Age}) + \gamma_3 (\text{Age} \times \text{Female})$$
. . .

where
$$\begin{aligned}
\gamma_0 &= \beta_0^\text{Male} 
&\gamma_1 &= \beta_0^\text{Female} - \beta_0^\text{Male} \\
\gamma_2 &= \beta_1^\text{Male} 
&\gamma_3 &= \beta_1^\text{Female} - \beta_1^\text{Male}
\end{aligned}$$

## Two models: Interaction in code

Generate data in 2019 that vary in both `sex` and `age`

```{r}
all_2019 <- simulated |>
  filter(year == 2019)
```
```{r, echo = F}
all_2019
```

## Two models: Interaction in code

. . .

The `*` operator allows slopes to differ across groups

```{r}
model <- lm(
  formula = income ~ sex * age,
  data = all_2019
)
```

```{r, echo = F, fig.height = 4}
p3
```

## Two models: Additive model in R

The `+` operator assumes slopes are the same across groups

```{r}
model <- lm(
  formula = income ~ sex + age,
  data = all_2019
)
```

```{r, echo = F, fig.height = 4}
target_subgroup <- tibble(age = 47, sex = c("female","male"))
target_subgroup$yhat <- predict(model, newdata = target_subgroup)

all_2019 |>
  mutate(yhat = predict(model)) |>
  group_by(sex, age) |>
  summarize_all(mean) |>
  ggplot(aes(x = age, y = income)) +
  geom_point(color = "gray", size = 1) +
  scale_y_continuous(
    name = "Annual Income Y",
    labels = label_currency(),
  ) +
  labs(x = "Age X") +
  facet_wrap(~sex) +
  # Regression line
  geom_line(aes(y = yhat)) +
  # Prediction
  geom_segment(
    data = target_subgroup,
    aes(x = 47, xend = 47, y = -Inf, yend = yhat), 
    color = "#2774AE", linetype = "dashed"
  ) +
  geom_segment(
    data = target_subgroup,
    aes(x = 47, xend = -Inf, y = yhat, yend = yhat), 
    color = "#2774AE", linetype = "dashed"
  ) +
  geom_point(
    data = target_subgroup,
    aes(x = age, y = yhat), 
    color = "#2774AE", size = 3
  )
```

## Interactions make lots of terms

```{r}
model <- lm(
  formula = income ~ sex * age * year,
  data = simulated
)
```

. . .

```{r}
summary(model)
```

## Interactions make lots of terms

```{r, echo = FALSE, fig.height = 4}
simulated |>
  mutate(yhat = predict(model)) |>
  group_by(sex, age, year) |>
  summarize_all(mean) |>
  ggplot(aes(x = age, y = income)) +
  geom_point(color = "gray", size = .7) +
  facet_grid(sex ~ year) +
  geom_line(aes(y = yhat)) +
  scale_y_continuous(
    name = "Income Y",
    labels = label_currency()
  ) +
  scale_x_continuous(breaks = c(35,45), name = "Age X")
```

# Penalized Regression

## Penalized regression

OLS is a linear model

$$\text{E}(Y\mid\vec{X}) = \beta_0 + \beta_1 X_1 + \beta_2 X_2 + \cdots$$

. . .

There are many linear models beyond OLS.

- (other ways of estimating the $\beta$ coefficients)

## Penalized regression

```{r, echo = FALSE}
simulate_female_2019 <- function(n) {
  read_csv("https://ilundberg.github.io/description/assets/truth.csv") |>
  filter(year == 2019 & sex == "female") |>
  slice_sample(n = n, weight_by = weight, replace = T) |>
  mutate(income = exp(rnorm(n(), meanlog, sdlog))) |>
  select(year, age, sex, income)
}
simulated <- simulate_female_2019(n = 100)

fit_models <- function(data, lambda = 5e3) {
  # Mean-center X and y
  # For simplicity, X is centered at 39.89634
  # which is its population mean in this simulation.
  X <- data$age - 39.89634
  y <- data$income - mean(data$income)
  ridge <- solve(t(X) %*% X + lambda, t(X) %*% y)
  ols <- coef(lm(y ~ -1 + X))
  newX <- 47 - 39.89634
  return(
    tibble(
      model = c("ols","ridge"),
      # Return intercept with X centered at 0
      intercept = mean(data$income) - 39.89634 * c(ols, ridge),
      slope = c(ols, ridge),
      yhat = c(
        mean(data$income) + newX * ols,
        mean(data$income) + newX * ridge
      )
    )
  )
}
results <- foreach(rep = 1:50, .combine = "rbind") %do% {
  simulate_female_2019(n = 100) |>
    fit_models(lambda = 1e4) |>
    mutate(rep = rep)
}
results |>
  mutate(model = ifelse(model == "ols", "Unpenalized","Penalized")) |>
  ggplot() +
  geom_segment(
    aes(x = 30, xend = 50,
        y = intercept + 30 * slope,
        yend = intercept + 50 * slope,
        group = rep),
    alpha = .8, color = "gray"
  ) +
  facet_wrap(~model) +
  scale_y_continuous(
    labels = label_currency()
  ) +
  labs(
    x = "Age", y = "Income", caption = "Among female respondents in 2019"
  )

```

## Penalized regression

```{r, echo = FALSE}
results |>
  mutate(model = ifelse(model == "ols", "Unpenalized","Penalized")) |>
  ggplot(aes(x = model, y = yhat)) +
  geom_jitter(width = .1, height = 0) +
  scale_y_continuous(labels = label_currency()) +
  labs(
    y = "Predicted mean income\namong 47-year-olds",
    x = "Model",
    caption = "Each dot is an estimate on a different\nsample from the population"
  )
```

## Unpenalized regression: In math

OLS chose $\alpha, \vec\beta$ to minimize this function:
$$
\begin{aligned}
\underbrace{\sum_i\left(Y_i - \hat{Y}_i\right)^2}_\text{Sum of Squared Error}
\end{aligned}
$$
where $\hat{Y}_i = \hat\alpha + \sum_j X_j \hat\beta_j$

## Penalized regression: In math

Penalized (ridge) regression chose $\alpha, \vec\beta$ to minimize this function:
$$
\begin{aligned}
\underbrace{\sum_i\left(Y_i - \hat{Y}_i\right)^2}_\text{Sum of Squared Error} + \underbrace{\lambda \sum_{j} \beta_j^2}_\text{Penalty Term}
\end{aligned}
$$
where $\hat{Y}_i = \hat\alpha + \sum_j X_j \hat\beta_j$

## Penalized regression: Code

```{r}
simulated <- simulate(n = 1e5)
```

## Penalized regression: Code

The `glmnet` package supports penalized regression
```{r}
library(glmnet)
```

## Penalized regression: Code

Create a model matrix of predictors

- Each column will correspond to a coefficient

```{r}
X <- model.matrix(~ age * sex * year, data = simulated)
```

. . .

Create a vector of the outcomes
```{r}
y <- simulated |> pull(income)
```

## Penalized regression: Code

Use the `cv.glmnet` function
```{r}
penalized <- cv.glmnet(
  x = X,    # model matrix we created
  y = y,    # outcome vector we created
  alpha = 0 # penalize sum of beta ^ 2
)
```

## Penalized regression: Code

```{r}
yhat <- predict(
  penalized,
  newx = X
)
```
```{r}
summary(yhat)
```

## When to use penalized regression?

. . .

- Many predictors and few observations
     - High-variance estimates
     
. . .
     
- When you are willing to accept bias
     - Model will be a bit wrong on average

## Linear regression: Learning goals

Some things you may know

- How to fit a linear model
- How to make predictions

Data science ideas

- Why model at all?
- Penalized linear regression