---
title: "Problem Set 3: Causal Inference"
---

**Due: 5pm on Friday, February 14.**

::: {.callout-note}
Want to see how you'll be evaluated? Check out the [rubric](https://docs.google.com/forms/d/e/1FAIpQLSed5nDlpx4vVcDu9Tz_2AFphUU4_ugLAw2EyF5h6124gT-bQQ/viewform?usp=sf_link)
:::

Student identifer: [type your anonymous identifier here]

- Use this [.qmd template](../assets/pset3/pset3.qmd) to complete the problem set
- In Canvas, you will upload the PDF produced by your .qmd file
- Put your identifier above, not your name! We want anonymous grading to be possible

## 1. Potential outcomes in a hypothetical setting

Jose says that coming to UCLA caused him to discover sociology, and it became his major! He says that if he had gone to UCSD, he would have stuck with biology.

**1.1 (5 points)**

In Jose's claim, what is the treatment?

**Answer.** Your answer here

**1.2 (5 points)**

Using the mathematical notation we discussed in class, define the two potential outcomes to which Jose is referring.

**Answer.** Your answer here

**1.3 (5 points) **

In a sentence or two, say how the Fundamental Problem of Causal Inference applies to Jose's claim.

**Answer.** Your answer here

**1.4 (5 points) **

Using conditional expectations or probabilities, write the following in math: the probability of majoring in sociology is higher among students who attend UCLA than among students who attend UCSD.

**Answer.** Your answer here

**1.5 (5 points) **

Using conditional expectations or probabilities, write the following in math: the probability of majoring in sociology would be higher if we intervened to send students to UCLA than if we intervened to send them to UCSD.

**Answer.** Your answer here

## 2. Potential outcomes in a real experiment

This problem is based on:

Bertrand, M \& Mullainathan, S. 2004. "[Are Emily and Greg More Employable Than Lakisha and Jamal? A Field Experiment on Labor Market Discrimination](https://www-jstor-org.proxy.library.cornell.edu/stable/3592802)." American Economic Review 94(4):991--1013.

Read the first 10 pages of the paper (through the end of section 2). In this paper,

- the unit of analysis is a resume submitted to a job opening
- the treatment is the name at the top of the resume
- the outcome is whether the employer called or emailed back for an interview

**2.1. (5 points) Fundamental Problem **

One submitted resume had the name "Emily Baker." It yielded a callback. The same resume could have had the name "Lakisha Washington." Explain how the Fundamental Problem of Causal Inference applies to this case (1--2 sentences).

**2.2. (5 points) Exchangeability **

In a sentence, what is the exchangeability assumption in this study? For concreteness, for this question you may suppose that the only names in the study were "Emily Baker" and "Lakisha Washington." Be sure to explicitly state the treatment and the potential outcomes.

**2.3 (15 points) Analyzing the experimental data **

Load packages that our code will use.
```{r, comment = F, message = F}
library(tidyverse)
library(haven)
```

Download the study's data from OpenICPSR: [https://www.openicpsr.org/openicpsr/project/116023/version/V1/view](https://www.openicpsr.org/openicpsr/project/116023/version/V1/view). This will require creating an account and agreeing to terms for using the data ethically. Put the data in the folder on your computer where this .Rmd is located. Read the data into R using `read_dta`. 

```{r, eval = F}
d <- read_dta("lakisha_aer.dta")
```
```{r, include = F}
d <- read_dta("../assets/data/lakisha_aer_aggregated.dta")
```

> If you have an error, you might need to set your working directory first. This tells R where to look for data files. At the top of RStudio, click Session -> Set Working Directory -> To Source File Location.

You will now see `d` in your Global Environment at the top right of RStudio.

We will use two variables:

| Name | Role | Values
| ---- | ---- | ------
`call` | outcome | 1 if resume submission yielded a callback
| | | 0 if not
`race` | category of treatments | `b` if first name signals Black
| | | `w` if first name signals white

The top of Table 1 reports callback rates: 9.65\% for white names and 6.45\% for Black names. Reproduce those numbers. Write code that reproduces these numbers.

