---
title: "Basics of R"
---

Now that you have installed R, it is time to try using R for the first time. Because there already exist many excellent resources on this topic, we will walk together in class through the material in [R4DS Ch 2](https://r4ds.hadley.nz/workflow-basics.html). This material will prepare you to complete Problem Set 1.

After the basics in [R4DS Ch 2](https://r4ds.hadley.nz/workflow-basics.html), return to this page and continue below to learn about an object we will often use in this course: a data frame.

## Using a data frame

The class of object we will use in this course most often is a `data.frame` or `tibble`, which is a particular type of data frame. A data frame is an object like the rectangular spreadsheet below.

| Address | Income | Number of People |
|---------|-----|-----|
| 5462 Park St | 54,896 | 2 |
| 4596 Ocean Ave Apt B | 22,465 | 1 |
| 6831 River Dr | 134,297 | 4 |

: A hypothetical data table with three households

The code below will load this data into an object in R.

```{r, echo = F, warning = F, message = F}
library(tidyverse)
basicData <- tibble(
  address = c(
    "5462 Park St",
    "4596 Ocean Ave Apt B",
    "6831 River Dr"
    ),
  income = c(54896,22465,134297),
  numPeople = c(2,1,4)
)
write_csv(basicData, file = "../data/basicData.csv")
```

```{r, eval = F}
library(tidyverse)
basicData <- read_csv("https://soc114.github.io/data/basicData.csv")
```

If you type `basciData` in your R console, you will see a printout of the data.

## Data frames and empirical questions

The structure of these data correspond to elements of empirical questions we study.

* The **outcome** is a variable to be summarized. It appears as a column in the data. Here, the outcome might be `income`.
* The **unit of analysis** is the unit for which the outcome is defined. Each unit corresponds to a row of the data. Here, the unit of analysis is a household.

The data may also contain other columns, such as the address of the household in this example. The other columns may represent **predictor variables** or variables that define **population subgroups**.

Now that you understand a data frame object, we will next use one to produce a visualization.


