---
title: "Linear Regression"
format: 
  html:
    fig-height: 3
---

```{r, echo = FALSE, message = FALSE, warning = FALSE}
library(tidyverse)
library(scales)
library(foreach)
options(pillar.print_max = 3, pillar.print_min = 3)
theme_set(theme_classic())
#update_theme(text = element_text(size = 18))
set.seed(90095)
```

> Here are slides in [website](../slides/lec05_linear_regression/lec05_linear_regression.qmd) and [pdf](../slides/lec05_linear_regression/lec05_linear_regression.pdf) format.

As an example, we will work with U.S. adult income by sex (male, female), age (30--50), and year (2010--2019). We will focus on the target population of those working 35+ hours per week for 50+ weeks per year. Data are simulated based on the 2010–2019 American Community Survey (ACS).

The function below will simulate data

```{r}
simulate <- function(n = 100) {
  read_csv("https://ilundberg.github.io/description/assets/truth.csv") |>
    slice_sample(n = n, weight_by = weight, replace = T) |>
    mutate(income = exp(rnorm(n(), meanlog, sdlog))) |>
    select(year, age, sex, income)
}
```

Below you can see this function in action.

```{r}
simulated <- simulate(n = 3e4)
```
```{r, echo = F}
simulated
```

## Conditional expectation

A key goal with linear regression is the conditional expectation: the mean of an outcome within a population subgroup.

- **expectation** refers to taking a mean
- **conditional** refers to within a subgroup

Example: Mean income among females age 47 in 2019 

Suppose we want to estimate that conditional mean in our data. One way is to first create the subgroup and take the mean among people in that subgroup in our sample.

`filter()` restricts our data to cases meeting requirements:

- the `sex` variable equals the value `female`
- the `age` variable equals the value `47`
- the `year` variable equals the value `2019`

```{r}
subgroup <- simulated |>
  filter(sex == "female") |>
  filter(age == 47) |>
  filter(year == 2019)
```

`summarize()` aggregates to the mean

```{r}
subgroup |>
  summarize(conditional_expectation = mean(income))
```

Often, we want to study many conditional expectations: the mean outcome in many subgroups. In previous homework, we have seen how `group_by` and `summarize` can yield the mean in many subgroups.

```{r}
simulated |>
  group_by(sex, age, year) |>
  summarize(conditional_expectation = mean(income))
```

In math, the **conditional expectation function** is the subgroup mean of $Y$ within a subgroup with the predictor values $\vec{X} = \vec{x}$. We use $\text{E}$ to denote the expectation operator. For simplicity, we will let $f()$ refer to the conditional expectation function, which has input $\vec{x}$ and outputs a conditional mean among those with $\vec{X} = \vec{x}$.

$$
f(\vec{x}) = \text{E}(Y\mid\vec{X} = \vec{x})
$$

To learn $f(\vec{x})$ from data is a central task in statistical learning.

#$ Statistical learning by pooling information

A common problem of statistical inference is that we want to study a subgroup, but there are few cases within the subgroup. For example, female respondents age 47 in 2019 in our simulated data.

```{r}
simulated |>
  filter(sex == "female") |>
  filter(year == 2019) |>
  filter(age == 47)
```

Few cases in a subgroup leads to statistical uncertainty about the mean in the subgroup. How can we better estimate this conditional mean?

One strategy is to **pool information** using a model. We have many female respondents in 2019. The only problem is that few are age 47.

```{r}
simulated |>
  filter(sex == "female") |>
  filter(year == 2019)
```

We might think that these other respondents (e.g., those age 46 and 48) are informative about the outcomes of 47-year-olds. There are many ways to pool information. A linear regression model is one strategy that pools information across people of all ages, to estimate a conditional mean at any particular age.

```{r, echo = FALSE, fig.height = 4}
female_2019 <- simulated |>
  filter(sex == "female") |>
  filter(year == 2019)
fit <- lm(income ~ age, data = female_2019)
target_yhat <- predict(fit, newdata = tibble(age = 47))
p <- female_2019 |>
  mutate(yhat = predict(fit)) |>
  group_by(sex, age) |>
  summarize_all(mean) |>
  ggplot(aes(x = age, y = income)) +
  geom_point(color = "gray") +
  scale_y_continuous(
    name = "Annual Income Y",
    labels = label_currency(),
  ) +
  labs(x = "Age X")
```

```{r, echo = F, fig.height = 4}
p <- p +
  # Regression line
  geom_line(aes(y = yhat)) +
  annotate(
    geom = "text", x = 37, y = 50e3, 
    label = "E('Y | X = x') == beta[0] + beta[1] * x",
    parse = TRUE, size = 5
  ) 
```

```{r, echo = F, fig.height = 4}
p <- p +
  # Prediction
  geom_segment(
    x = 47, xend = 47, y = -Inf, yend = target_yhat, 
    color = "#2774AE", linetype = "dashed"
  ) +
  annotate(geom = "text", x = 47.5, y = 45e3, label = "X = 47", color = "#2774AE", hjust = 0, size = 5) +
  geom_point(
    x = 47, y = target_yhat, 
    color = "#2774AE", size = 3
  ) +
  geom_segment(
    x = 47, xend = -Inf, y = target_yhat, yend = target_yhat, 
    color = "#2774AE", linetype = "dashed"
  ) +
  annotate(
    geom = "text", x = 40, y = target_yhat, 
    label = "hat(Y) == ~hat(E)('Y | X' == 47) == ~hat(beta)[0] + hat(beta)[1]%*%47", 
    parse = TRUE, size = 5,
    color = "#2774AE",
    vjust = -1
  )
p
```

The model assumes that all of the conditional means fall along a line. Then, it estimates the intercept $\beta_0$ and slope $\beta_1$ of this line to best fit those conditional means. Finally, one can use the model to make a prediction at any particular $X$ value.

### Practice question

$$
\text{E}(Y\mid X) = \beta_0 + \beta_1 X
$$

Suppose $\beta_0 = 5$ and $\beta_1 = 3$

1. What is the conditional mean when $X = 0$?
2. What is the conditional mean when $X = 1$?
3. What is the conditional mean when $X = 2$?
4. How much does the conditional mean change for each unit increase in $X$?

## Coding a linear model

Coding a linear model in R is easy. First, generate some data (get the `simulate()` function from futher up this page).
```{r}
simulated <- simulate(n = 3e4)
```

For our example, restrict to female respondents in 2019.
```{r}
female_2019 <- simulated |>
  filter(sex == "female") |>
  filter(year == 2019)
```

Then learn a model from the data with the `lm()` function.

```{r}
model <- lm(
  formula = income ~ age, 
  data = female_2019
)
```

Here is how that code worked:

- `model` is an object of class `lm` for **l**inear **m**odel
- `lm()` function creates this object
- `formula` argument is a model formula
     - `outcome ~ predictor` is the syntax
- `data` is a dataset containing `outcome` and `predictor`

We can look at the learned model with the `summary()` function.

```{r}
summary(model)
```

Finally, we might predict at a new X value. First, define the data at which to make the prediction: a person age 47.
```{r}
to_predict <- tibble(age = 47)
```

Predict for that subgroup
```{r}
predict(model, newdata = to_predict)
```

To review, our model **pooled information**:

- People of all ages contributed to `model`
- Then we predicted at a single age

### Practice question

Below is the line fit to the population data. Suppose we want to learn $\text{E}(\log(Y)\mid X = 30)$.

```{r, echo = FALSE, fig.height = 3}
read_csv("https://ilundberg.github.io/description/assets/truth.csv") |>
  filter(sex == "female" & year == 2019) |>
  ggplot(aes(x = age, y = meanlog)) +
  geom_point() +
  geom_smooth(method = "lm", se = F) +
  labs(
    y = "Mean of log income",
    x = "Age",
    caption = "Among female respondents in 2019"
  )
```

1. Why might this model make a misleading estimate?
2. Why might the model still be useful?

## Additive vs interactive models

Below, we visualize two models: one for male and one for female respondents.

```{r, echo = FALSE, fig.height = 4}
all_2019 <- simulated |>
  filter(year == 2019)
model <- lm(income ~ age*sex, data = all_2019)
target_subgroup <- tibble(age = 47, sex = c("female","male"))
target_subgroup$yhat <- predict(model, newdata = target_subgroup)

p <- all_2019 |>
  mutate(yhat = predict(model)) |>
  group_by(sex, age) |>
  summarize_all(mean) |>
  ggplot(aes(x = age, y = income)) +
  geom_point(color = "gray", size = 1) +
  scale_y_continuous(
    name = "Annual Income Y",
    labels = label_currency(),
  ) +
  labs(x = "Age X") +
  facet_wrap(~sex)
```
```{r, echo = F, fig.height = 4}
p2 <- p +
  # Regression line
  geom_line(aes(y = yhat)) +
  geom_text(
    data = tibble(
      sex = c("female","male"),
      age = 30, income = 51e3,
      label = c(
        "E('Y | X = x, Sex = Female') == {beta[0]}^'Female' + {beta[1]}^'Female' * x",
        "E('Y | X = x, Sex = Male') == {beta[0]}^'Male' + {beta[1]}^'Male' * x"
      )
    ),
    aes(label = label),
    parse = T, size = 2, hjust = 0
  )
```
```{r, echo = F, fig.height = 4}
p3 <- p2 +
  # Prediction
  geom_segment(
    data = target_subgroup,
    aes(x = 47, xend = 47, y = -Inf, yend = yhat), 
    color = "#2774AE", linetype = "dashed"
  ) +
  geom_segment(
    data = target_subgroup,
    aes(x = 47, xend = -Inf, y = yhat, yend = yhat), 
    color = "#2774AE", linetype = "dashed"
  ) +
  geom_point(
    data = target_subgroup,
    aes(x = age, y = yhat), 
    color = "#2774AE", size = 3
  )
p3
```

There are two equivalent ways to describe these two models. The first is by thinking of them as separate linear regressions.

$$
\begin{aligned}
\text{E}(Y\mid X, \text{Female}) &= \beta_0^\text{Female} + \beta_1^\text{Female}\times \text{Age} \\
\text{E}(Y\mid X, \text{Male}) &= \beta_0^\text{Male} + \beta_1^\text{Male}\times \text{Age} \\
\end{aligned}
$$

The second way is to think of them as one pooled linear regression that **interacts** age and sex to allow the slope on age to differ by sex.

$$\text{E}(Y \mid X, \text{Sex}) = \gamma_0 + \gamma_1(\text{Female}) + \gamma_2(\text{Age}) + \gamma_3 (\text{Age} \times \text{Female})$$

Note that both approaches summarize the conditional mean function with 4 parameters. The two approaches are actually equivalent, as you can show with some algebra.

$$\begin{aligned}
\gamma_0 &= \beta_0^\text{Male} 
&\gamma_1 &= \beta_0^\text{Female} - \beta_0^\text{Male} \\
\gamma_2 &= \beta_1^\text{Male} 
&\gamma_3 &= \beta_1^\text{Female} - \beta_1^\text{Male}
\end{aligned}$$

Below, we practice writing an interaction in code. First generate data in 2019 that vary in both `sex` and `age`.

```{r}
all_2019 <- simulated |>
  filter(year == 2019)
```
```{r, echo = F}
all_2019
```

The `*` operator allows slopes to differ across groups

```{r}
model <- lm(
  formula = income ~ sex * age,
  data = all_2019
)
```

```{r, echo = F, fig.height = 4}
p3
```

## Two models: Additive model in R

The `+` operator assumes slopes are the same across groups

```{r}
model <- lm(
  formula = income ~ sex + age,
  data = all_2019
)
```

```{r, echo = F, fig.height = 4}
target_subgroup <- tibble(age = 47, sex = c("female","male"))
target_subgroup$yhat <- predict(model, newdata = target_subgroup)

all_2019 |>
  mutate(yhat = predict(model)) |>
  group_by(sex, age) |>
  summarize_all(mean) |>
  ggplot(aes(x = age, y = income)) +
  geom_point(color = "gray", size = 1) +
  scale_y_continuous(
    name = "Annual Income Y",
    labels = label_currency(),
  ) +
  labs(x = "Age X") +
  facet_wrap(~sex) +
  # Regression line
  geom_line(aes(y = yhat)) +
  # Prediction
  geom_segment(
    data = target_subgroup,
    aes(x = 47, xend = 47, y = -Inf, yend = yhat), 
    color = "#2774AE", linetype = "dashed"
  ) +
  geom_segment(
    data = target_subgroup,
    aes(x = 47, xend = -Inf, y = yhat, yend = yhat), 
    color = "#2774AE", linetype = "dashed"
  ) +
  geom_point(
    data = target_subgroup,
    aes(x = age, y = yhat), 
    color = "#2774AE", size = 3
  )
```

When you have many interactions, the model starts to have lots of terms! This can make interpretation hard. But, you can always use the model to predict any conditional mean you want, even if there are many interactions.

```{r}
model <- lm(
  formula = income ~ sex * age * year,
  data = simulated
)
```

```{r}
summary(model)
```

The many terms of the model correspond to many slopes in subgroups, as visualized below.

```{r, echo = FALSE, fig.height = 4}
simulated |>
  mutate(yhat = predict(model)) |>
  group_by(sex, age, year) |>
  summarize_all(mean) |>
  ggplot(aes(x = age, y = income)) +
  geom_point(color = "gray", size = .7) +
  facet_grid(sex ~ year) +
  geom_line(aes(y = yhat)) +
  scale_y_continuous(
    name = "Income Y",
    labels = label_currency()
  ) +
  scale_x_continuous(breaks = c(35,45), name = "Age X")
```

## Penalized Regression

We close with a more advanced data science topic: penalized regression. We will discuss a method known as ridge regression or L2 penalized regression. This data science approach is a linear model just like OLS, but estimates the coefficients slightly differently.

Before defining this type of penalized regression, below we show what happens when we use it. Each line is a regression estimated on a different sample from the population. What similarities and differences do you notice between the penalized and unpenalized regressions?

```{r, echo = FALSE, message = FALSE, warning = FALSE}
simulate_female_2019 <- function(n) {
  read_csv("https://ilundberg.github.io/description/assets/truth.csv") |>
  filter(year == 2019 & sex == "female") |>
  slice_sample(n = n, weight_by = weight, replace = T) |>
  mutate(income = exp(rnorm(n(), meanlog, sdlog))) |>
  select(year, age, sex, income)
}
simulated <- simulate_female_2019(n = 100)

fit_models <- function(data, lambda = 5e3) {
  # Mean-center X and y
  # For simplicity, X is centered at 39.89634
  # which is its population mean in this simulation.
  X <- data$age - 39.89634
  y <- data$income - mean(data$income)
  ridge <- solve(t(X) %*% X + lambda, t(X) %*% y)
  ols <- coef(lm(y ~ -1 + X))
  newX <- 47 - 39.89634
  return(
    tibble(
      model = c("ols","ridge"),
      # Return intercept with X centered at 0
      intercept = mean(data$income) - 39.89634 * c(ols, ridge),
      slope = c(ols, ridge),
      yhat = c(
        mean(data$income) + newX * ols,
        mean(data$income) + newX * ridge
      )
    )
  )
}
results <- foreach(rep = 1:50, .combine = "rbind") %do% {
  simulate_female_2019(n = 100) |>
    fit_models(lambda = 1e4) |>
    mutate(rep = rep)
}
results |>
  mutate(model = ifelse(model == "ols", "Unpenalized","Penalized")) |>
  ggplot() +
  geom_segment(
    aes(x = 30, xend = 50,
        y = intercept + 30 * slope,
        yend = intercept + 50 * slope,
        group = rep),
    alpha = .8, color = "gray"
  ) +
  facet_wrap(~model) +
  scale_y_continuous(
    labels = label_currency()
  ) +
  labs(
    x = "Age", y = "Income", caption = "Among female respondents in 2019"
  )

```

We could use each approach to predict the mean income among 47-year-olds. Below are those predictions, with one dot from each simulated sample. How do the patterns below align with what you noticed in the graph above?

```{r, echo = FALSE}
results |>
  mutate(model = ifelse(model == "ols", "Unpenalized","Penalized")) |>
  ggplot(aes(x = model, y = yhat)) +
  geom_jitter(width = .1, height = 0) +
  scale_y_continuous(labels = label_currency()) +
  labs(
    y = "Predicted mean income\namong 47-year-olds",
    x = "Model",
    caption = "Each dot is an estimate on a different\nsample from the population"
  )
```

The reason one uses penalized regression is to reduce the sampling variance of estimates, at the cost of some bias (estimates are generally drawn in toward the overall sample mean). To understand how, it is useful to see some math.

OLS chose $\alpha, \vec\beta$ to minimize this function:
$$
\begin{aligned}
\underbrace{\sum_i\left(Y_i - \hat{Y}_i\right)^2}_\text{Sum of Squared Error}
\end{aligned}
$$
where $\hat{Y}_i = \hat\alpha + \sum_j X_j \hat\beta_j$

Penalized (ridge) regression chose $\alpha, \vec\beta$ to minimize this function:
$$
\begin{aligned}
\underbrace{\sum_i\left(Y_i - \hat{Y}_i\right)^2}_\text{Sum of Squared Error} + \underbrace{\lambda \sum_{j} \beta_j^2}_\text{Penalty Term}
\end{aligned}
$$
where $\hat{Y}_i = \hat\alpha + \sum_j X_j \hat\beta_j$

The only difference between the two is that penalized regression seeks to avoid having a large value of $\sum_j \beta_j^2$. Thus, it prefers to estimate models with coefficients near zero. In practice, researchers often mean-center covariates and outcomes for penalized regression so that this pulls all estimates toward the overall mean.

### Penalized regression in code

Penalized regression is available through many R packages. Here we illustrate with the `glmnet` package. First, simulate a large sample.

```{r, message = FALSE, warning = FALSE}
simulated <- simulate(n = 1e5)
```

Load the package.
```{r, message = FALSE, warning = FALSE}
library(glmnet)
```

Create a model matrix of predictors

- This converts the predictor data into matrix form
- Each column will correspond to a coefficient

```{r}
X <- model.matrix(~ age * sex * year, data = simulated)
```

Create a vector of the outcomes
```{r}
y <- simulated |> pull(income)
```

Use the `cv.glmnet` function to call the package. For now, we will leave as a black box how it chooses the penalty parameter $\lambda$.

```{r}
penalized <- cv.glmnet(
  x = X,    # model matrix we created
  y = y,    # outcome vector we created
  alpha = 0 # penalize sum of beta ^ 2
)
```

Finally, make predictions from the model at each observed data point.

```{r}
yhat <- predict(
  penalized,
  newx = X
)
```
```{r}
summary(yhat)
```

When should you use penalized regression? The key reasons are motivated by the original illustration figures above. You should use penalized regression to reduce the variance of your estimates. This may occur in settings where you have many predictors and few observations, for example. However, there is a cost: penalized regression generally yields biased estimates of conditional means, so the model will be wrong on average.

In future classes, we will discuss data-driven ways to choose among the many statistical and machine-learning approaches to estimate conditional mean functions.